---
title: "Genus level analyses"
author: "Sven Kleine Bardenhorst"
date: "`r format(Sys.time(), '%d %B %Y')`"
mail: "s.kleinebardenhorst@uni-muenster.de"
twitter: "r_graph_gallery"
github: "svenkb"
home: "epi.uni-muenster.de"
# !!! You need to provide a logo image here !!! Or just delete the field for no logo
logo: "Logo_WWU.svg"
output:
  epuRate::epurate:
    toc: TRUE
    number_sections: FALSE
    code_folding: "hide"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE,fig.width = 12)
```

```{css include=FALSE}
body .main-container {
max-width: 1600px;
}
```



```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

```{r load packages, message=FALSE, warning=FALSE, include=FALSE}
#### load packages ####
source('packages.R')

#### load functions ####
source("functions.R")

#### Level
level <- "Genus"
```

# Beta diversity 

## Prepare full distance matrix

```{r echo=FALSE, message=FALSE}
dat <- readRDS("phyl_list.RDS")

beta <- prepare_beta_data(dat,tax=level,method = "bray")
```

```{r}
cmdscale(beta$dist,k=2) %>%
  as.data.frame() %>%
  rownames_to_column("ID") %>%
  dplyr::left_join(beta$meta) %>%
  ggplot(aes(x=V1,y=V2,col=author,shape=status)) +
  geom_point()

```



```{r message=FALSE, warning=FALSE, include=FALSE}
beta_NMDS <- metaMDS(as.matrix(beta$dist,labels = T),distance = "bray",
                     k=2,
                     maxit = 999)
```

```{r}
stressplot(beta_NMDS)
```


```{r}
fit <- adonis(beta$dist~status,data = beta$meta,strata = beta$meta$author,permutations = 9999)
summary(fit)
```

```{r}
fit1 <- betadisper(beta$dist,as.factor(beta$meta$author))

disper_d <- data.frame(distance=fit1$distances,group=fit1$group)

ggplot(data=disper_d,aes(x=group,y=distance,fill=group)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=90),
        axis.title.x = element_text("Study"))


```

# Differential abundance

## Differential abundance by permutation

We implemented a non-parametric test based on permutations, as it is - compared to negative binomial models - robust against violations of assumptions of parametric models. We adopted the test statistic as provided by Thorsen et al. as follows

$$
\begin{equation}
T = log(\frac{\bar{\theta}_{PD}}{\bar{\theta}_{HC}})
\end{equation}
$$

where $\bar{\theta}$ is the mean relative abundance in the respective study group.

```{r}
full.dat <- prepare_DA_data(dat,tax=level)
full.dat <- full.dat %>% dplyr::select(ID,status,author,N,everything())
```


```{r message=FALSE, warning=FALSE, include=FALSE}
otu <- full.dat %>% dplyr::select(-c(1:3)) %>% mutate_all(function(x) x/.$N) %>% dplyr::select(-N)
meta <- full.dat %>% dplyr::select(c(1:4))

group <- meta$status
strata <- meta$author



cores <- detectCores()-1
cl <- makeCluster(cores)
clusterEvalQ(cl, library("permute"))

perm.out <- parApply(cl=cl,otu,2,perm_DA,group=group,type="mean",n.perm=9999,strata=strata,plot=F)

stopCluster(cl)

perm.p <- data.frame(names=names(perm.out),matrix(unlist(perm.out),nrow=length(perm.out),byrow=T))
colnames(perm.p) <- c("Genera","p.value","log.mean.diff")


perm.p$q.value <- p.adjust(perm.p$p.value,method = "fdr")
perm.p <- perm.p %>% mutate(mean.ratio = round(exp(log.mean.diff),2)) %>% arrange(q.value)

#perm.p %>% filter(Genera == "Prevotella")

#full.dat %>%
#  group_by(author,status) %>%
#  summarize(abundance=sum(Marvinbryantia),
#            prevalence=sum(Marvinbryantia >1))


#orig <- c("Bifidobacterium","Bacteroides","Lactobacillus","Blautia","Coprococcus",
#              "Dorea","Roseburia","Faecalibacterium","Oscillospira","Akkermansia",
#              "Acinetobacter","Alistipes","Anaerotruncus","Aquabacterium","Butyricicoccus",
#              "Citrobacter","Clostridium.IV","Clostridium.XlVa","Clostridium.XlVb","Clostridium.XVIII",
#              "Desulfovibrio","Enterococcus","Fusicatenibacter","Gordonibacter","Holdemania",
#              "Klebsiella","Lachnospiracea_incertae_sedis","Lactococcus","Parabacteroides","Paraprevotella",
#              "Peptoniphilus","Prevotella","Ruminococcus","Salmonella","Sediminibacterium","Shigella",
#              "Sphingomonas")
```



The table below lists the genera which are significant after FDR-correction  and - for comparison - the genera which were reported in the original publications.

```{r}
reported <- readxl::read_xlsx(".//reported_results.xlsx") %>% filter(!is.na(genus)) %>% mutate(Genera = genus) %>% dplyr::select(Genera,reported,increased,decreased) 


perm.res <- perm.p %>%
  left_join(reported,by = "Genera") %>%
  replace(is.na(.),0) %>%
  filter(p.value < 0.05| reported > 0)


perm.res %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))


```

# Negative binomial mixed models

```{r message=FALSE, warning=FALSE}
taxa <- names(full.dat)[-c(1:4)] # Extract all taxa names


cores <- detectCores()-1
cl <- makeCluster(cores)
clusterEvalQ(cl, library("NBZIMM"))

univ.fit <- parSapply(cl=cl,taxa,fitZINBMM,full.dat,simplify = F)

stopCluster(cl)
comp.error <- taxa[sapply(univ.fit,is.null)]

#fit.sig <- univ.fit[unlist(lapply(univ.fit[!sapply(univ.fit,is.null)],is.sig))]

negbin.res <- round(t(sapply(univ.fit[!sapply(univ.fit,is.null)],summarize_univariate)),5) %>%
  data.frame() %>%
  rownames_to_column("Genera") %>%
  mutate(q.value = p.adjust(p.value,method="fdr")) %>%
  dplyr::select(Genera,estimate,p.value,q.value) %>%
  filter(p.value < 0.05) %>%
  left_join(reported,by = "Genera") %>%
  replace(is.na(.),0) %>%
  filter(p.value < 0.05 | reported > 0) %>%
  arrange(desc(reported),estimate)

negbin.res %>% kable() %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

# Summarized results

The following table shows all results which were significant using both methods, as well as all previously reported results.
```{r echo=FALSE, fig.height=10}
all.res <- perm.res %>%
  full_join(negbin.res,by="Genera",suffix = c(".perm", ".nb")) %>%
  dplyr::select(Genera,mean.ratio,q.value.perm,estimate,q.value.nb,reported.perm,increased.perm,decreased.perm) %>%
  dplyr::rename(reported = reported.perm,
                increased = increased.perm,
                decreased = decreased.perm,
                estimate.nb = estimate) %>%
  filter(q.value.perm < 0.05 & q.value.nb < 0.05 | reported > 0) %>% arrange(desc(reported))

all.res %>% kable %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


```{r fig.height=300}
scientific_10 <- function(x) {
  parse(text=gsub("e", " %*% 10^", scales::scientific_format()(x)))
}


sig <- all.res$Genera

meta %>% bind_cols(otu) %>% dplyr::select(status,author,all_of(sig)) %>% melt %>% mutate(points=ifelse(value==0,NA,value)) %>%#  group_by(author,status,variable) %>% summarize_all(mean) %>%
  ggplot(aes(x=author,y=value,fill=status),data=.) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(fill = status,y=points), size = 1, shape = 21, position = position_jitterdodge()) +
  #geom_bar(stat="identity",position="dodge") +
  theme(axis.text.x = element_text(angle=90)) +
  xlab("Relative abundance") +
  scale_y_continuous(label=scientific_10) +
  facet_wrap(~variable,ncol=1,scales = "free") 

```

