---
title: "Phylum level analyses"
author: "Sven Kleine Bardenhorst"
date: "`r format(Sys.time(), '%d %B %Y')`"
mail: "s.kleinebardenhorst@uni-muenster.de"
twitter: "r_graph_gallery"
github: "svenkb"
home: "epi.uni-muenster.de"
# !!! You need to provide a logo image here !!! Or just delete the field for no logo
logo: "Logo_WWU.svg"
output:
  epuRate::epurate:
    toc: TRUE
    number_sections: FALSE
    code_folding: "hide"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE,fig.width = 12,cache = T)
```

```{css include=FALSE}
body .main-container {
max-width: 1600px;
}
```



```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

```{r load packages, message=FALSE, warning=FALSE, include=FALSE}
#### load packages ####
source('packages.R')

#### load functions ####
source("functions.R")

#### Level
level <- "Phylum"
```

# Beta diversity 

## Prepare full distance matrix

```{r echo=FALSE, message=FALSE}
dat <- readRDS("phyl_list.RDS")

beta <- prepare_beta_data(dat,tax=level,method = "bray")
```

```{r}
cmdscale(beta$dist,k=2) %>%
  as.data.frame() %>%
  rownames_to_column("ID") %>%
  dplyr::left_join(beta$meta) %>%
  ggplot(aes(x=V1,y=V2,col=author,shape=status)) +
  geom_point()

```



```{r message=FALSE, warning=FALSE, include=FALSE}
beta_NMDS <- metaMDS(as.matrix(beta$dist,labels = T),distance = "bray",
                     k=2,
                     maxit = 999)
```

```{r}
stressplot(beta_NMDS)
```


```{r}
fit <- adonis(beta$dist~status,data = beta$meta,strata = beta$meta$author,permutations = 9999)
summary(fit)
```

```{r}
fit1 <- betadisper(beta$dist,as.factor(beta$meta$author))

disper_d <- data.frame(distance=fit1$distances,group=fit1$group)

ggplot(data=disper_d,aes(x=group,y=distance,fill=group)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=90),
        axis.title.x = element_text("Study"))


```

# Differential abundance

## Differential abundance by permutation

We implemented a non-parametric test based on permutations, as it is - compared to negative binomial models - robust against violations of assumptions of parametric models. We adopted the test statistic as provided by Thorsen et al. as follows

$$
\begin{equation}
T = log(\frac{\bar{\theta}_{PD}}{\bar{\theta}_{HC}})
\end{equation}
$$

where $\bar{\theta}$ is the mean relative abundance in the respective study group.

```{r}
full.dat <- prepare_DA_data(dat,tax=level)
full.dat <- full.dat %>% dplyr::select(ID,status,author,N,everything())
```


```{r message=FALSE, warning=FALSE, include=FALSE}
otu <- full.dat %>% dplyr::select(-c(1:3)) %>% mutate_all(function(x) x/.$N) %>% dplyr::select(-N)
meta <- full.dat %>% dplyr::select(c(1:4))

group <- meta$status
strata <- meta$author

cores <- detectCores()-1
cl <- makeCluster(cores)
clusterEvalQ(cl, library("permute"))

perm.out <- parApply(cl=cl,otu,2,perm_DA,group=group,type="mean",n.perm=9999,strata=strata,plot=F)

stopCluster(cl)

perm.p <- data.frame(names=names(perm.out),matrix(unlist(perm.out),nrow=length(perm.out),byrow=T))
colnames(perm.p) <- c("Phyla","p.value","log.mean.diff")


perm.p$q.value <- p.adjust(perm.p$p.value,method = "fdr")
perm.p <- perm.p %>% mutate(mean.ratio = round(exp(log.mean.diff),2)) %>% arrange(q.value)

#perm.p %>% filter(Genera == "Prevotella")

#full.dat %>%
#  group_by(author,status) %>%
#  summarize(abundance=sum(Marvinbryantia),
#            prevalence=sum(Marvinbryantia >1))


#orig <- c("Bifidobacterium","Bacteroides","Lactobacillus","Blautia","Coprococcus",
#              "Dorea","Roseburia","Faecalibacterium","Oscillospira","Akkermansia",
#              "Acinetobacter","Alistipes","Anaerotruncus","Aquabacterium","Butyricicoccus",
#              "Citrobacter","Clostridium.IV","Clostridium.XlVa","Clostridium.XlVb","Clostridium.XVIII",
#              "Desulfovibrio","Enterococcus","Fusicatenibacter","Gordonibacter","Holdemania",
#              "Klebsiella","Lachnospiracea_incertae_sedis","Lactococcus","Parabacteroides","Paraprevotella",
#              "Peptoniphilus","Prevotella","Ruminococcus","Salmonella","Sediminibacterium","Shigella",
#              "Sphingomonas")
```



The table below lists the Phyla which are significant after FDR-correction  and - for comparison - the Phyla which were reported in the original publications.

```{r}
reported <- readxl::read_xlsx(".//reported_results.xlsx") %>%
  filter(!is.na(phylum) & is.na(class) & is.na(order) &  is.na(family) & is.na(genus)) %>%
  mutate(Phyla = phylum) %>%
  dplyr::select(Phyla,reported,increased,decreased)


perm.res <- perm.p %>%
  left_join(reported,by = "Phyla") %>%
  replace(is.na(.),0) %>%
  filter(p.value < 0.05| reported > 0) %>%
  arrange(desc(reported))


perm.res %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))


```

# Negative binomial mixed models

```{r message=FALSE, warning=FALSE}
taxa <- names(full.dat)[-c(1:4)] # Extract all taxa names


cores <- detectCores()-1
cl <- makeCluster(cores)
clusterEvalQ(cl, library("NBZIMM"))

univ.fit <- parSapply(cl=cl,taxa,fitZINBMM,full.dat,simplify = F)

stopCluster(cl)
comp.error <- taxa[sapply(univ.fit,is.null)]

#fit.sig <- univ.fit[unlist(lapply(univ.fit[!sapply(univ.fit,is.null)],is.sig))]

negbin.res <- round(t(sapply(univ.fit[!sapply(univ.fit,is.null)],summarize_univariate)),5) %>%
  data.frame() %>%
  rownames_to_column("Phyla") %>%
  mutate(q.value = p.adjust(p.value,method="fdr")) %>%
  dplyr::select(Phyla,estimate,p.value,q.value) %>%
  filter(p.value < 0.05) %>%
  left_join(reported,by = "Phyla") %>%
  replace(is.na(.),0) %>%
  filter(p.value < 0.05 | reported > 0) %>%
  arrange(desc(reported),estimate)

negbin.res %>% kable() %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

# Summarized results

The following table shows all results which were significant using both methods, as well as all previously reported results.
```{r echo=FALSE, fig.height=10}
all.res <- perm.res %>%
  full_join(negbin.res,by="Phyla",suffix = c(".perm", ".nb")) %>%
  dplyr::select(Phyla,mean.ratio,q.value.perm,estimate,q.value.nb,reported.perm,increased.perm,decreased.perm) %>%
  dplyr::rename(reported = reported.perm,
                increased = increased.perm,
                decreased = decreased.perm,
                estimate.nb = estimate) %>%
  filter(q.value.perm < 0.05 & q.value.nb < 0.05 | reported > 0) %>% arrange(desc(reported))

all.res %>% kable %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


```{r fig.height=50}
scientific_10 <- function(x) {
  parse(text=gsub("e", " %*% 10^", scales::scientific_format()(x)))
}


sig <- all.res$Phyla

meta %>% bind_cols(otu) %>% dplyr::select(status,author,all_of(sig)) %>% melt %>% mutate(points=ifelse(value==0,NA,value)) %>%#  group_by(author,status,variable) %>% summarize_all(mean) %>%
  ggplot(aes(x=author,y=value,fill=status),data=.) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(fill = status,y=points), size = 1, shape = 21, position = position_jitterdodge()) +
  #geom_bar(stat="identity",position="dodge") +
  theme(axis.text.x = element_text(angle=90)) +
  xlab("Relative abundance") +
  scale_y_continuous(label=scientific_10) +
  facet_wrap(~variable,ncol=1,scales = "free") 

```

# Stratified analyses

```{r}
full.dat <- prepare_DA_data(dat,tax=level)
glimpse(full.dat)
library(meta)
library(broom)

# Augment tidy function with new variables to extract



tidy_n <- function(x) {
  tidydf = tidy(x, conf.int=T)
  tidydf$N = length(x$residuals)
  tidydf$Outcome = colnames(x$model)[1]
  tidydf
}

#fit <- (glm.nb(Firmicutes~status+offset(log(N)),data=full.dat))
#summary(fit)

#colnames(fit$model)[1]

# Univariate models on nested datasets

#meta_df <- full.dat %>%
#  group_by(author) %>%
#  nest() %>%
#  mutate(model = map(data, function(x) glm.nb(Firmicutes~status+offset(log(N)),data=x)),
#         est = map(model,tidy_n)) %>%
#  dplyr::select(author,est) %>%
#  unnest() %>%
#  filter(term=="statusPD") %>%
#  mutate(p.value = round(p.value,4),
#         TE = estimate,
#         seTE = std.error,
#         .after = author,
#         .keep = "unused")


# Meta-analysis with random effect 

#m <- meta::metagen(TE,
#             seTE,
#             data=meta_df,
#             studlab=paste(author),
#             outclab = "Firmicutes",
#             comb.fixed = F,
#             comb.random = T,
#             prediction=TRUE,
#             sm="RR")

# Make forest plot


#meta::forest(m)




# Multivariate apply, to apply univariate models per taxon per study

# Select taxa names
taxa <- setdiff(colnames(full.dat),c("ID","status","author","N"))



# Prepare model wrapper to iterate over all bacteria -  with tryCatch to deal with possible non-convergence
negbin_model <- function(.x){
  sapply(taxa,function(x) {
    tryCatch(glm.nb(as.formula(paste0(x,"~status+offset(log(N))")),data=.x),error=function(e) print(paste0("Something went wrong!: ",e," ", x)))
    }
    ,simplify = F)
}


mdf <- full.dat %>%
  group_by(author) %>%
  nest() %>%
  mutate(model = map(.x = data,negbin_model)) %>% 
  unnest(model)


meta_model <- function(x) {

meta::metagen(TE,
             seTE,
             data=x,
             studlab=paste(author),
             comb.fixed = F,
             comb.random = T,
             prediction=TRUE,
             outclab =x$Outcome,
             sm="RR")
}


is.converged <- function(x) {
  !coef(summary(x))["statusPD",2] > 10000

}

meta_models <- mdf %>%
  filter(!map_lgl(model,is.character )) %>%
  filter(map_lgl(model,is.converged)) %>%
  mutate(est = map(model,tidy_n)) %>%
  dplyr::select(author,est) %>%
  unnest() %>%
  filter(term=="statusPD") %>%
  mutate(p.value = round(p.value,4),
         TE = (estimate),
         seTE = (std.error),
         .after = author,
         .keep = "unused") %>%
  arrange(Outcome) %>%
  ungroup() %>% group_by(Outcome) %>%
  nest() %>%
  mutate(model = map(data,meta_model))


```



```{r fig.height=30, fig.width=10}

library(ggthemr)
library(firatheme)
library(tvthemes)
library(bbplot)
library(ggpomological)
library(ggtech)
library(hrbrthemes)
library(ggalt)

tidyMetagen <- function(model,
                        parameter = c("studlab","TE","seTE","lower","upper","statistic","pval","df","TE.fixed",
                                      "seTE.fixed","lower.fixed","upper.fixed","zval.fixed","statistic.fixed","pval.fixed",
                                      "TE.random","seTE.random","lower.random","upper.random","zval.random","statistic.random",
                                      "pval.random","null.effect","seTE.predict","lower.predict","upper.predict","level.predict",
                                      "w.random","w.fixed",
                                      "I2")) {
  
  out <- reduce(model[parameter],data.frame,check.rows=F)
  colnames(out) <- parameter
  
  return(out)
}


getWeights <- function(x) x/sum(x)*100


plot_df <- meta_models %>%
  mutate(est = map(model,tidyMetagen)) %>%
  dplyr::select(Outcome,est) %>%
  unnest %>%
  mutate(TE = round(exp(TE),2),
         seTE = round(exp(seTE),2),
         upper = round(exp(upper),2),
         lower = round(exp(lower),2),
         pval.random = round(pval.random,2),
         pval.fixed = round(pval.fixed,2)) %>%
  group_by(Outcome) %>%
  mutate(weights.r = round(getWeights(w.random),1),
         weights.f = round(getWeights(w.fixed),1)) %>% ungroup


meta_effects <- plot_df %>%
  group_by(Outcome) %>%
  slice(1) %>%
  ungroup() %>%
  dplyr::select(Outcome,TE.fixed:pval.random) %>%
  pivot_longer(TE.fixed:upper.random) %>%
  mutate(estimates = gsub(".fixed","",gsub(".random","",name)),
         names = sub('.*\\.', '', name),.keep="unused") %>%
  pivot_wider(values_from = value,
              names_from = estimates) %>%
  mutate(studlab = names,
         TE = round(exp(TE),2),
         seTE = round(exp(seTE),2),
         upper = round(exp(upper),2),
         lower = round(exp(lower),2),
         weights.r = 100,
         weights.f = 100,
         .keep="unused")

final_df <- plot_df %>%
  
  bind_rows(meta_effects) %>%
  group_by(Outcome) %>%
  mutate(I2 = round(mean(I2,na.rm=T),2)*100) %>%
  ungroup %>%
  mutate(studlab = relevel(factor(studlab),"random")) %>%
  filter(studlab != "fixed") %>%
  arrange(Outcome,studlab) %>%
  mutate(fullab = case_when(studlab == "random" ~ paste0(TE," [",lower,"; ",upper,"] ",weights.r," %","\n I² = ",I2," % ; p = ",pval.random),
                            studlab != "random" ~ paste0(TE," [",lower,"; ",upper,"] ",weights.r," %")),
         studlab =  plyr::revalue(studlab,c("random"="Random effect model")),
         pval.rand = pval.random,
         pval.fix = pval.fixed) %>%
  dplyr::select(-ends_with(".fixed")) %>%
  dplyr::select(-ends_with(".random")) %>%
  dplyr::select(-ends_with(".predict")) %>%
  filter(Outcome != "Tenericutes") %>%
  filter(Outcome != "Fusobacteria") %>%
  filter(Outcome != "Elusimicrobia") %>%
  filter(pval.rand < 0.05)

## TO DO

# Get weights
# Get I^2
# get tau^2

levels(final_df$studlab)


te_color <- "darkgrey"
overall_color <- "darkred"

pal <- c(
  "Random effect model" = overall_color,
  "fixed" = overall_color,
  "Aho et al., 2019" = te_color, 
  "Barichella et al., 2019" = te_color, 
  "Heintz-Buschart et al., 2018" = te_color,
  "Keshavarzian et al., 2015" = te_color, 
  "Pietrucci et al., 2019" = te_color, 
  "Qian et al., 2018" = te_color, 
  "Weis et al., 2019" = te_color
)



p <- ggplot(aes(y=studlab,x=TE,color=studlab),data=final_df) +
  #geom_linerange(aes(xmin=lower,xmax=upper,width=.1),col="blue") +
  #geom_errorbar(aes(xmin=lower,xmax=upper,width=.25),size=1.25) +
  geom_dumbbell(aes(x=lower,xend=upper),size=1.3,size_xend=1.5,size_x=1.5,colour_x="black",colour_xend="black") + 
  geom_point(shape=18,size=3,color="black") +
  scale_x_log10(limits=c(-15,55000),breaks=c(.01,.1,.33,3,10,100,500),labels = c(.01,.1,.33,3,10,100,"")) +
  facet_grid(Outcome~.,scales = "free",space="free") +
  xlab("Incidence Rate Ratio") +
  geom_vline(xintercept = 1, linetype="dashed",size=1,col="black") + 
  #theme_classic()+
  theme_ipsum(base_family = "Times New Roman",
              grid = "X",
              axis_title_size = 12,axis_title_just = "cc") +
  #theme_hrbrmstr() + 
  geom_text(aes(label=as.character(final_df$fullab)),y=final_df$studlab,x=4,size=3) +
  ylab(NULL) +
  scale_colour_manual(values = pal) +
  theme(legend.position = "none") +
  ggtitle("Two-stage meta analyses")

p

ggsave("Plots/meta_phylum.PDF",plot=p,device = cairo_pdf,dpi=300,width=200,height = 290,units="mm")

```


## Effects of harmonization

```{r}
dat <- readRDS("phyl_list.RDS")
dat_fs <- readRDS("phyl_list_FS.RDS")
level <- "Phylum"
df <- prepare_DA_data(dat,tax=level)
df_fs <- prepare_DA_data(dat_fs,tax=level)


df_fs[df_fs$ID=="SRR8351974","Gemmiger"]

otu_table(dat[[1]])
otu_table(dat_fs[[1]])
df_fs$Gemmiger == df$Gemmiger

tax_table(dat[[1]])[3,]
tax_table(dat_fs[[1]])[3,]



library(meta)
library(broom)


df$status

fit <- glm(Firmicutes~status,data=df,family=quasipoisson())
fit0 <- glm(Firmicutes~1,data=df,family=quasipoisson())


anova(fit,fit0,test="F")



summar <- summary(fit)

coef(summary(fit))

```












## Bayesian alternative
```{r eval=FALSE, include=FALSE}

fit <- brms::brm(data = full.dat, family = negbinomial,
    as.formula(paste0("Firmicutes","~status+offset(log(N))")),
    prior = c(prior(normal(0, 10), class = Intercept),
              prior(normal(0, 1), class = b),
              prior(gamma(0.01, 0.01), class = shape)),  # this is the brms default
    iter = 4000, warmup = 1000, cores = 2, chains = 2,
    seed = 11)


str(fit)
```





